# Reproducible Research


## Report submitted towards Assignment 01

###Part 01: Environment Setup
Getting the environment ready with required packages
```{r, echo = TRUE}
# Assuming the packages are already installed
library(ggplot2)
library(tidyr)
library(lubridate)
library(dplyr)
library(knitr)
```


###Part 02: Loading and preprocessing the data
Load the data off the csv file and process the fields in desired format
```{r, echo = TRUE}
# Assuming the data doesn't needs to be downloaded
# Read the data off the csv file. Change the data frame to data frame table and convert date to POSIXct
data <- read.csv("activity.csv", header = TRUE)
data <- tbl_df(data)
data <- mutate(data, date = as.POSIXct(date))
```

###Part 03: Total number of steps taken per day?
Create aggregation (sum of  Steps by date). Plot it on a graph and reports the mean/median
```{r, echo = TRUE}

aggdata <- aggregate(data$steps, by = list(data$date), FUN = sum, na.rm = TRUE)
hist(aggdata$x, xlab = "# of Steps", ylab = "# of Days", main = "Histogram # of Steps")
abline(v = mean(aggdata$x), col = "red", lwd = "2")
abline(v = median(aggdata$x), col = "green", lwd = "2")
legend("topright", c('Mean','Median'),lwd=c(2,2),col=c('red','green'))
```

The mean of total number of steps taken per day (red) `r as.integer(mean(aggdata$x))` and the median (green) is `r as.integer(median(aggdata$x))`


###Part 04: Average daily activity pattern
Create aggregation (sum of  Steps by time-interval), plot it on a graph and reports the mean/median
```{r echo = TRUE}
minutedata <- aggregate(data$steps, by = list(data$interval), FUN = mean, na.rm = TRUE)
colnames(minutedata) <- c("Grp", "avgSteps")
plot(minutedata$Grp, minutedata$avgSteps, type = 'l', lwd = "2", col = 'blue', xlab = "Time-Interval (Minute) of the Day", ylab = "Average Steps", main = "Average Daily Activity Pattern")
abline(v = minutedata[minutedata$avgSteps == max(minutedata$avgSteps),1], col = "green", lwd = "2")
legend("topright", c('Steps','Maximum Steps time interval'), lwd = c(2,2), col = c('blue','green'))
```

The time interval with maxium number of steps is `r minutedata[minutedata$avgSteps == max(minutedata$avgSteps),1]`

###Part 05: Imputing missing values
There are a total of `r as.integer(count(data[data$steps == NA,]))` missing values in the data. Therefore it is critical to detemine and impute the missing values. The strategy is to replace any missing values with the mean of corresponding 5-minute time interval.

```{r echo = TRUE}
# Create a copy of original data. Convert that to data frame table
data2 <- data
data2 <- tbl_df(data2)
minutedata <- tbl_df(minutedata)

#megre it with the minute data created in earlier step and replicate missing values with corresponsing interval average
data2 <- merge(data2,minutedata, by.x = "interval", by.y = "Grp")
data2 <- mutate(data2, stepsII = ifelse(is.na(steps),avgSteps,steps))

# New aggregation is created using the new dataset (with no missing values) by date and a graph is generated
aggdata2 <- aggregate(data2$stepsII, by = list(data$date), FUN = sum, na.rm = TRUE)
hist(aggdata2$x, xlab = "# of Steps", ylab = "# of Days", main = "Histogram # of Steps")
abline(v = mean(aggdata2$x), col = "red", lwd = "2")
abline(v = median(aggdata2$x), col = "green", lwd = "2")
legend("topright", c('Mean','Median'),lwd=c(2,2),col=c('red','green'))
```

The mean of total number of steps taken per day (red) and the median (green)  after imputing the missng values is `r as.integer(mean(aggdata2$x))`  and `r as.integer(median(aggdata2$x))` respectively. The new values (after imputing) have a mean higher by `r (as.integer(mean(aggdata2$x)) - as.integer(mean(aggdata$x)))` and median higher by `r (as.integer(median(aggdata2$x)) - as.integer(median(aggdata$x)))` steps.

### Part 06: Differences in activity patterns between weekdays and weekends
```{r echo = TRUE}
# A new variable is created to identfy between the weekday or the weekend
data2 <- mutate(data2, dayofweek = weekdays(date, abbreviate = TRUE))
data2 <- mutate(data2, Wkday = as.factor(ifelse(dayofweek %in% c("Mon","Tue","Wed","Thu","Fri"), "weekday", "weekend")))

# Number of steps are plotted to compare trends between weekend and weekdays steps
minutedata2 <- aggregate(data2$stepsII, by = list(data2$interval, data2$Wkday), FUN = mean, na.rm = TRUE)
colnames(minutedata2) <- c("Interval","Weekday","avgSteps")
qplot(Interval, avgSteps, data = minutedata2, geom = c("line"), facets = Weekday~.)+ggtitle("Average Steps by 5 Minute time interval")+labs(x="Interval",y="Average Steps")

<<<<<<< HEAD
```
=======
```
>>>>>>> 1df932737a6c174628070f66f9192857f4f44041
